---
title: "census_var"
format: html
author: ElectriGrid
warning: false
message: false
---


This quarto document explores the Census ACS 5-year data utilizing the `tidycensus` package. The challenge is to narrow down the data from the 13,033 variables. The final output of this repository is a clean data frame to be utilized in the ElectriGrid MEDS capstone. The client has provided a list of variables of interest. 

 - B25003_001: housing_units_tenure
 - B25003_002: owner_occupied
 - B25003_003: renter_occupied
 - B25003H_003: renter_occupied_white # tenure
 - B25026_001: pop_in_occupied # total pop in occupied units
 - B25026_009: pop_renter_occupied
 - B25127_001: total_housing_units
 - B11011_004: married_one_unit # married-couple family, one-unit structures; single family housing rate
 - B11011_002: household_type_by_unit # family households/household type by units in structure
 - B16010_041: pop_bach_or_higher
 - B16010_001: total_pop
 - B25077_001: median_value # dollars
 - B25097_001: mortgage_status # dollars
 - B25107_001: median_value_by_year # value by year structure built
 - B19113_001: median_family_income_2015 # past 12 months; 2015 inflation-adjusted
 - B19013_001: median_household_income # past 12 months; 2015 inflation-adjusted
 - B19013A_001: median_family_income_white # past 12 months; 2018 inflation-adjusted,
 - B19013B_001: median_family_income_black # past 12 months; 2018 inflation-adjusted
 - B01001A_001: sex_by_age_white
 - B01001B_001: sex_by_age_black
 - B01001C_001: sex_by_age_american_indian
 - B01001D_001: sex_by_age_asian
 - B01001E_001: sex_by_age_hawaian_and_pacific_islander
 - B01001F_001: sex_by_other_race
 - B01001G_001: sex_by_

```{r}
# load necessary libraries
library(tidyverse)
library(tidycensus)
library(janitor)
```

# Census variable

Download a csv with the variables for the census data.

```{r}
# access variable names to for selection of 
var20 <- load_variables(2020, "acs5", cache = TRUE)

#........................write data to csv.......................
write_csv(var20, file = here::here("census-data", "var20.csv")) 
```

There are slightly under 30,000 variables.

#### Clean the labels

Once the labels column is cleaned it can be utilized to filter down the variables. Here we keep only the numbers and letters.

```{r}
var20_clean <- var20 |> 
  # clean label column
  mutate(label = str_replace_all(label, "[^A-Za-z0-9]", " ")) 
```

There is no argument for the level of variables when downloading the csv so we'll filter down to census tract. This is also specified when calling data but let's subset here to see the true number of variables change with future filtering.

```{r}
tract_var20 <- var20_clean |> filter(geography == "tract")
```

We only need some of the census data not all 13,000 rows.

This loop can be applied to all data frames created before.

The loop is just outputting one of variables' values but we need all of them. We can just make a list of the variables and get the estimates for all of them in one call. This will make it easier to get a tidy data frame of the variables and the estimates we need.

```{r}
socioeconomic_var <- c("B25003_001", 
                       "B25003_002", 
                       "B25003_003", 
                       "B25003H_003", 
                       "B25026_001", 
                       "B25026_009", 
                       "B25127_001",
                       "B11011_004",
                       "B11011_002",
                       "B16010_041", 
                       "B16010_001",
                       "B25077_001", 
                       "B25097_001",
                       "B25107_001",  
                       "B19113_001",
                       "B19013_001", 
                       "B19013A_001", 
                       "B19013B_001", 
)
```


```{r}
tidy_df <- df |>
  select(GEOID, NAME, variable, estimate, geometry) |> 
 pivot_wider(names_from = variable, values_from = c(estimate)) 

#........................write data to csv.......................
#write_csv(pop_census_tidy, file = here::here("census-data", "pop_census_tidy.csv"))  
```
